The function--sheaf dictionary for smooth commutative group schemes over finite fields, with applications over local fields

### 2014-08-21 dictionary.tex
Gone through Section 3,4 and 5, again.

Throughout this paper, $G$ is a smooth, commutative group scheme
over a finite field $\Fq$, and $m : \GxG\to G$ is its multiplication morphism.

@@

Let $\Frob{}$ denote the geometric Frobenius element in $\Gal(\bFq/\Fq)$ as
well as the corresponding automorphism of $\Spec{\bFq}$. The Weil group
$\Weil{}\subset \Gal(\bFq/\Fq)$ is the subgroup generated by $\Frob{}$.
Let $\Frob{G} \ceq \id_{G} \times \Frob{}$ be the Frobenius automorphism of $\bG$.
%When no other fields play a role we will write $\Frob{}$ and $\Weil{}$ instead.

@@

\subsection{Objects, when $G$ is etale}%\texorpdfstring{$S$ and $T$}{S and T}}

@@

\section{Quasicharacter sheaves on commutative algebraic groups}

%In this section we find $\QCiso{G}$ and $\Aut(\qcs{L})$ for commutative algebraic group over $\Fq$; note that we found these in Section~\ref{sec:etale} when $G$ is etale.

%\subsection{Bounded quasicharacter sheaves}\label{sec:bounded}
\subsection{Descent to G}\label{sec:bounded}

%Let $G$ be a smooth commutative group scheme over $\Fq$.
In this section we consider a category of sheaves on $G$ obtained by
replacing the Weil sheaf $(\gqcs{L}, \phi)$ on $\bG$ in the definition of a quasicharacter sheaf with an $\ell$-adic local system on $G$ itself.

\begin{definition}
Let $\bQC(G)$ be the category of pairs $(\qcs{E},\mu_\qcs{E})$
where $\qcs{E}$ an $\ell$-adic local system on $G$ of rank~one,
equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes \qcs{E}$
satisfying the analogue of \ref{QC.1} on $G$;
morphisms in $\bQC(G)$ are defined as in the second part of
\ref{QC.3}.
%This is the category of \emph{bounded quasicharacter sheaves} on $G$.
\end{definition}

@@

We will identify $\bQC(G)$ with its image under $B_G$ and say that a quasicharacter sheaf $\qcs{L} \in \QC(G)$ \emph{descends to $G$} if it is isomorphic to some $B_G(\qcs{E}, \mu_\qcs{E})$.
%We will see in Corollary~\ref{cor:bounded-and-finite} that bounded quasicharacter sheaves are exactly quasicharacter sheaves with bounded Trace of Frobenius.\todo{Check this, later.}

@@

%Here we describe a class of quasicharacter sheaves on $G$ that will play a role in Section~\ref{sec:restriction}.

@@

We have just seen that, for a connected commutative algebraic group $G$ over $\Fq$, the category of quasicharacters $\qcs{L}$ on $G$ is equivalent to the category of characters $\psi$ of $G(\Fq)$.
We have also just seen if the character of $\Lang^*\qcs{L}$ is $\psi$ then the canonical isomorphism $m^*\psi \iso \psi \boxtimes \psi$ determines the isomorphism $\mu : \qcs{L} \to \qcs{L}\boxtimes\qcs{L}$ which is part of the data of the quasicharacter sheaf $\qcs{L}$. That leads to a perspective on the function--sheaf dictionary common in the literature in which one considers $1$-dimensional local systems $\qcs{L}$ on $G$ for which \emph{there exists} an isomorphism $m^*\qcs{L} \iso \qcs{L} \boxtimes\qcs{L}$ \cite{kamgarpour:09a}*{Introduction}, or sometimes, $1$-dimensional local systems $\gqcs{L}$ on $\bG$ for which \emph{there exists} an isomorphism $\Frob{G}^*\gqcs{L} \iso \gqcs{L}$ and an isomorphism $\bm^*\gqcs{L} \iso \gqcs{L} \boxtimes\gqcs{L}$.

%In the introduction to this paper we noted that, when working with connected commutative algebraic groups over finite fields, Deligne's dictionary can be described using Weil sheaves $\qcs{L}$ for which there exists an isomorphism $m^* \qcs{L} \iso \qcs{L} \boxtimes\qcs{L}$, without keeping track of the isomorphism.

Although the category of quasicharacter sheaves on commutative smooth group schemes $G$ over finite fields does specialize to that common notion when $G$ is connected and of finite type, that notion is not appropriate if one wishes to extend the dictionary to commutative smooth group schemes over finite fields. 
In particular, as we have already seen in Section~\ref{sec:etale}, for a given quasicharacter sheaf $\qcs{L} = (\gqcs{L},\mu,\phi)$ there may be a different isomorphism $\mu' : \bm^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$ for which $\qcs{L}'=(\gqcs{L},\mu',\phi)$ is also a quasicharacter sheaf on $G$, not isomorphic to $\qcs{L}$. 
Indeed, Proposition~\ref{prop:etale-iso} shows that $\Hh^0(\Weil{},\Hh^2(\bG,\EEx))$ measures the possibilities for such $\mu'$ when $G$ is etale.
We will see later in this section that, for any smooth commutative group scheme $G$, the possibilities for such $\mu'$ are measured by $\Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx))$.

@@

Consider the short exact sequence
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is a etale commutative group scheme, and this smooth,
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$
in Sections~\ref{sec:connected} and \ref{sec:etale}, respectively.
In this section we show that \eqref{eq:pi2} is right exact.

%We say that a morphism $f : H\to G$ of commutative group $\Fq$-schemes is a {\it discrete isogeny} to $G$ if it is a finite, surjective \'etale morphism and
the action of $\Gal(\bFq/\Fq)$ on the \'etale group scheme $\ker f$ is trivial.

@@

\begin{proof}
Let $\pi: B \to G^0$ be a discrete isogeny and set $A \ceq \ker \pi$.
  We will find a discrete isogeny $f: H\to G$
  such that that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.  Namely, we will fit $\pi$ into the following diagram,
  \begin{equation}\label{extension-diagram}
  \begin{tikzcd}
  A \arrow[equal]{r} \dar & A \dar \\
  B \rar \dar[swap]{\pi} & H \rar \dar[swap]{f} & \pi_0(H) \arrow{d}[below,rotate=90]{\sim}[swap]{\pi_0(f)} \\
  G^0 \rar & G \rar & \pi_0(G),
  \end{tikzcd}
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.

@@

  We therefore have the existence of diagram \eqref{extension-diagram}
  at the level of $\bFq$-points.  This expresses $H(\bFq)$ as a
  disjoint union of translates of $B(\bFq)$; by transport of structure
  we may take $H$ to be a group scheme over $\bFq$.  Similarly, the
  restriction of $f$ to each component of $H$ is a morphism of
  schemes, and thus $f$ is as well.  Finally, the whole diagram
  descends to a diagram of $\Fq$-schemes since the $\bFq$-points of
  the objects come equipped with continuous $\Gamma$-actions and the
  morphisms are $\Gamma$-equivariant.
\end{proof}

In order to use the results of Section~\ref{sec:connected}, we need to know that
identity component of $G$ is actually an algebraic group.

\begin{lemma} \label{lem:G0alg-grp}
If $G$ is a commutative smooth group scheme over $\Fq$ then $G^0$ is a connected algebraic group.
\end{lemma}
\begin{proof}
 Since $G$ is a smooth group scheme over $\Fq$, its
 identity component $G^0$ of $G$ is a connected smooth,
 group scheme of finite type over $\Fq$, reduced over some finite extension of $\Fq$
 \cite{vdGeer-Moonen:AbelianVarieties}*{3.17}.
 Since $\Fq$ is a finite field and hence perfect, $G^0$ is actually reduced over $\Fq$
 \cite{EGAIV2}*{Prop 6.4.1}.  Since every group scheme over a field is separated
 \cite{vdGeer-Moonen:AbelianVarieties}*{3.12},
 it follows that $G^0$ is a connected algebraic group.
\end{proof}

@@

We saw in Proposition~\ref{prop:functorialG} that Trace of Frobenius
$\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is a functorial group homomorphism.
In this section we improve that result.

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme over $\Fq$ then
  \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
is surjective with kernel canonically isomorphic to $\Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$.
\end{theorem}

\begin{proof}
  Recall the short exact sequence \eqref{eq:pi0}
  defining the component group scheme for $G$:
  \[
  \begin{tikzcd}
    1 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
  \end{tikzcd}
  \]
  Let
  \[
  \begin{tikzcd}[row sep=20]
    0 \rar & \ker \TrFrob{\pi_0(G)} \arrow[dashed]{r} \dar & \arrow{d} \ker \TrFrob{G} \arrow[dashed]{r} & \ker \TrFrob{G^0} =0 \arrow[dotted, out=-10, in=170]{dddll} \dar & \\
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar \dar
    & \arrow{d} G(\Fq)^* \rar & G^0(\Fq)^* \rar \dar & 0\\
    & \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \coker \TrFrob{G} \arrow[dashed]{r} &  \coker \TrFrob{G^0} =0 &
  \end{tikzcd}
  \]
  be the commutative diagram of abelian groups obtained by applying
  Proposition~\ref{prop:pullback} to \eqref{eq:pi0}.
 %
  The sequence of abelian groups
  \begin{equation}\label{eq:pi0k}
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0,
  \end{tikzcd}
  \end{equation}
  is exact since $\Hh^1(\Fq,G^0) =0$ by Lemma~\ref{lem:G0alg-grp} and Lang's theorem on connected algebraic groups over finite fields \cite{lang:56a}.
  Since $\EEx$ is divisible, $\Hom_\text{Ab}(\ - \ ,\EEx)$ is exact and thus the dual sequence of
  character groups is also exact.
%
  The upper row is exact by Proposition~\ref{prop:middleexact}.
  Now Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}
  imply that $\ker \TrFrob{G^0} =0$ and $\coker \TrFrob{G^0}=0$,
  while Proposition~\ref{prop:etale-iso} gives $\ker \TrFrob{\pi_0(G)} \iso \Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$
  and $\coker \TrFrob{\pi_0(G)}=0$.
%
It follows from the snake lemma that $\coker \TrFrob{G} =0$
and $\ker \TrFrob{\pi_0(G)} \to \ker \TrFrob{G}$ is an isomorphism.
\end{proof}

\todo{Insert comments on $\Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$. 
It is trivial if $G$ is an algebraic group with cyclic component group.
Lusztig}

@@

\begin{proposition}\label{prop:basechange}
With notation above,
the rule $\nu_{k'/k}: (\gqcs{L}, \mu, \phi) \mapsto (\gqcs{L}, \mu, \phi_{k'})$
 defines a monoidal functor $\QC(G) \to \QC(G_{k'})$.
 Moreover, the following diagram commutes:
\[
\begin{tikzcd}[column sep=60]
\QCiso{G} \rar{\nu_{k'/k}} \dar{\TrFrob{G}} & \QCiso{G_{k'}} \dar{\TrFrob{G_{k'}}} \\
G(k)^*  \rar{\Nm^*} & G(k')^*.
\end{tikzcd}
\]
\end{proposition}

@@

Finally, let $G'$ be a smooth commutative group scheme over $k'$;
we explain how to geometrize the canonical isomorphism between characters of $G'(k')$ and of $(\Res_{k'/k}G')(k)$.
We may decompose the base change $(\Res_{k'/k}G')_{k'}$ of $\Res_{k'/k}G'$ to $k'$
into a product of copies of $G'$, indexed by elements of $\Gal(k'/k)$:
\[
(\Res_{k'/k}G')_{k'} \cong \prod_{\Gal(k'/k)} G'.
\]
Since products and coproducts agree for group schemes we have a natural inclusion of $k'$-schemes
\[
G' \hookrightarrow (\Res_{k'/k}G')_{k'}
\]
mapping $G'$ into the summand corresponding to $1 \in \Gal(k'/k)$.  Composing $\nu_{k'/k}$
from Proposition~\ref{prop:basechange} with pullback along this map yields a functor
\[
\rho : \QC(\Res_{k'/k}G') \to \QC(G').
\]

\begin{proposition}
Let $k'/k$ be a finite extension and let $G'$ be a smooth commutative group scheme over $k'$.
The following diagram of isomorphisms commutes, where the bottom map is the identity:
\[
\begin{tikzcd}
\QCiso{\Res_{k'/k} G'} \dar{\TrFrob{\Res_{k'/k} G'}} \rar{\rho} & \QCiso{G'} \dar{\TrFrob{G'}}\\
G'(k')^* \rar & G'(k')^*
\end{tikzcd}
\]
\end{proposition}
\begin{proof}
By Proposition~\ref{prop:pullback} the pullback part of the definition of $\rho$ corresponds to the map
\[
(\Res_{k'/k}G')(k') \to G'(k')^*
\]
induced by $g \mapsto (g, 1, \ldots, 1)$.  Since the action of $\Gal(k'/k)$ on
\[
(\Res_{k'/k}G')_{k'} \cong \prod_{\Gal(k'/k)} G'
\]
is given by permuting coordinates, composition with the norm map yields the identity on $G'(k')$.
\end{proof}

\todo{Can we improve on this result by proving that $\rho$
is an equivalence of categories?}

@@

Let $K$ be a local field (in the sense of Serre) with residue field $\Fq$. 
In this section we consider connected smooth commutative algebraic groups over $K$ that admit a Néron model, by which we means a lft Néron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected smooth commutative algebraic groups over $K$ that contain no subgroup of type $\mathbb{G}_\text{a}$.
These algebraic groups over $K$ form an additive category, denoted by $\mathcal{N}/K$ below, in a natural way, as a full subcategory of connected commutative algebraic groups over $K$. This category includes algebraic tori over $K$ and Abelian varieties over $K$.

Let $R$ be the ring of integers of $K$.
Write $\m$ for the maximal ideal of $R$ and set $R_n = R/\m^{n+1}$, for every non-negative integer $n$.
Suppose $X_K \in \mathcal{N}/K$.
A {\it quasicharacter of $X_K(K)$} is a group homomorphism $X_K(K) \to \EEx$ that factors through $X_K(X) \to X(R_n)$, for some non-negative integer $n$;
the group of quasicharacters of $X_K(K)$ will be denoted by $\Hom_\text{qc}(X_K(K),\EEx)$ and the subgroup of those that factor through $X(R_n)$ will be denoted by $\Hom_n(X_K(K),\EEx)$.
%
In this section we explain how to geometrize and categorify quasicharacters of $X_K(K)$ using quasicharacter sheaves.

@@

Let $K$, $R$ and $R_n$ be as above.
Set $S = \Spec{R}$.
For every non-negative integer $n$,  and $S_n = \Spec{R_n}$;  
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.

>>>>

Throughout this paper, $G$ is a smooth commutative group scheme
over a finite field $\Fq$, and $m : \GxG\to G$ is its multiplication morphism.

@@

Let $\Frob{}$ denote the geometric Frobenius element in $\Gal(\bFq/\Fq)$ as
well as the corresponding automorphism of $\Spec{\bFq}$. The Weil group
$\Weil{}\subset \Gal(\bFq/\Fq)$ is the subgroup generated by $\Frob{}$.
Let $\Frob{G} \ceq \id_{G} \times \Frob{}$ be the Frobenius automorphism of $\bG = G \times_{\Spec{\Fq}} \Spec{\bFq}$.
%When no other fields play a role we will write $\Frob{}$ and $\Weil{}$ instead.

@@

\subsection{Objects, when $G$ is etale}\label{sec:SandT}

@@

\section{Quasicharacter sheaves on connected commutative algebraic groups}

Before turning our attention to quasicharacter sheaves on connected commutative algebraic groups in Section~\ref{sec:connected}, we being with two sections that treat two special classes of quasicharacter sheaves on arbitrary smooth commutative group schemes $G$: those that descend from local systems on $\bG$ to local systems on $G$ in Section~\ref{sec:bounded}; and those local systems on $G$ that are defined by discrete isogenies onto $G$ in Section~\ref{sec:discrete_isogenies}. The latter will play a role in Sections~\ref{sec:connected} and \ref{sec:restriction}.

\subsection{Descent to G}\label{sec:bounded}

In this section we consider a category of sheaves on $G$ obtained by
replacing the Weil sheaf $(\gqcs{L}, \phi)$ on $\bG$ in the definition of a quasicharacter sheaf with an $\ell$-adic local system on $G$ itself.

\begin{definition}
Let $\bQC(G)$ be the category of pairs $(\qcs{E},\mu_\qcs{E})$
where $\qcs{E}$ an $\ell$-adic local system on $G$ of rank~one,
equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes \qcs{E}$
satisfying the analogue of \ref{QC.1} on $G$;
morphisms in $\bQC(G)$ are defined as in the second part of
\ref{QC.3}.
\end{definition}

@@

We will say that a quasicharacter sheaf $\qcs{L} \in \QC(G)$ \emph{descends to $G$} if it is isomorphic to some $B_G(\qcs{E}, \mu_\qcs{E})$.

@@

@@

We have just seen that, for a connected commutative algebraic group $G$ over $\Fq$, the category of quasicharacter sheaves on $G$ is equivalent to the category of characters $\psi$ of $G(\Fq)$.
We have also just seen that if the character of $\Lang^*\qcs{E}$ is $\psi$ then the canonical isomorphism $m^*\psi \iso \psi \boxtimes \psi$ determines the isomorphism $\mu_\qcs{E} : \qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ which is part of the data that determines a quasicharacter sheaf on $G$.
That leads to a perspective on the function--sheaf dictionary common in the literature in which one considers $1$-dimensional local systems $\qcs{E}$ on $G$ for which \emph{there exists} an isomorphism $m^*\qcs{E} \iso \qcs{E} \boxtimes\qcs{E}$.
% as in \cite{kamgarpour:09a}*{Introduction} (or sometimes, $1$-dimensional local systems $\gqcs{L}$ on $\bG$ for which \emph{there exists} an isomorphism $\Frob{G}^*\gqcs{L} \iso \gqcs{L}$ and an isomorphism $\bm^*\gqcs{L} \iso \gqcs{L} \boxtimes\gqcs{L}$).
%
Although the category of quasicharacter sheaves on $G$ does specialize to that common notion when $G$ is connected and of finite type, the common notion is not appropriate if one wishes to extend the dictionary to smooth commutative group schemes over finite fields. 
In particular, as we have already seen in Section~\ref{sec:etale}, for a given quasicharacter sheaf $\qcs{L} = (\gqcs{L},\mu,\phi)$ there may be a different isomorphism $\mu' : \bm^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$ for which $\qcs{L}'=(\gqcs{L},\mu',\phi)$ is also a quasicharacter sheaf on $G$, not isomorphic to $\qcs{L}$ in $\QC(G)$.
Indeed, Proposition~\ref{prop:etale-iso} shows that $\Hh^0(\Weil{},\Hh^2(\bG,\EEx))$ measures the possibilities for $\mu'$ when $G$ is etale.
We will see in Section~\ref{sec:main} that, for any smooth commutative group scheme $G$, the possibilities for $\mu'$ are measured by $\Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx))$.

@@

Consider the short exact sequence
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is a etale commutative group scheme -- and thus smooth --
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{\pi_0(G)}$ and $\QCiso{G^0}$
in Sections~\ref{sec:SandT} and \ref{sec:connected}, respectively.
We will shortly see that \eqref{eq:pi2} is exact.

@@

\begin{proof}
Let $\pi: B \to G^0$ be a discrete isogeny; set $A \ceq \ker \pi$.
  We will find a discrete isogeny $f: H\to G$
  such that that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.  Namely, we will fit $\pi$ into the following diagram,
  \begin{equation}\label{extension-diagram}
  \begin{tikzcd}
  A \arrow[equal]{r} \dar & A \dar \\
  B \rar \dar[swap]{\pi} & H \rar \dar[swap]{f} & \pi_0(H) \arrow{d}[below,rotate=90]{\sim}[swap]{\pi_0(f)} \\
  G^0 \rar & G \rar & \pi_0(G),
  \end{tikzcd}
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.

@@

  We therefore have the existence of diagram \eqref{extension-diagram}
  at the level of $\bFq$-points.  This expresses $H(\bFq)$ as a
  disjoint union of translates of $B(\bFq)$; by transport of structure
  we may take $H$ to be a group scheme over $\bFq$.  Similarly, the
  restriction of $f$ to each component of $H$ is a morphism of
  schemes, and thus $f$ is as well.  Finally, the whole diagram
  descends to a diagram of $\Fq$-schemes since the $\bFq$-points of
  the objects come equipped with continuous $\Gal(\bFq/\Fq)$-actions and the
  morphisms are $\Gal(\bFq/\Fq)$-equivariant.
\end{proof}

We now wish to use the apply the results of Section~\ref{sec:connected} to the identity component of $G$; to do so, we must know that the identity component of $G$ is actually an algebraic group over $\Fq$.

\begin{lemma} \label{lem:G0alg-grp}
If $G$ is a commutative smooth group scheme over $\Fq$ then its identity component,$G^0$ is a connected algebraic group over $\Fq$.
\end{lemma}
\begin{proof}
 Since $G$ is a smooth group scheme over $\Fq$, its
 identity component $G^0$ of $G$ is a connected smooth,
 group scheme of finite type over $\Fq$, reduced over some finite extension of $\Fq$
 \cite{vdGeer-Moonen:AbelianVarieties}*{3.17}.
 Since $\Fq$ is a finite field and hence perfect, $G^0$ is actually reduced over $\Fq$
 \cite{EGAIV2}*{Prop 6.4.1}.  Since every group scheme over a field is separated
 \cite{vdGeer-Moonen:AbelianVarieties}*{3.12},
 it follows that $G^0$ is a connected algebraic group.
\end{proof}

@@

We saw in Proposition~\ref{prop:functorialG} that $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is a functorial group homomorphism.
In this section we improve that result.

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth commutative group scheme over $\Fq$ then
  \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
is surjective with kernel canonically isomorphic to $\Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$.
\end{theorem}

\begin{proof}
  Recall the short exact sequence \eqref{eq:pi0}
  defining the component group scheme for $G$:
  \[
  \begin{tikzcd}
    1 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
  \end{tikzcd}
  \]
  Let
  \[
  \begin{tikzcd}[row sep=20]
  & 0 \arrow{d} & 0 \arrow{d} & 0\arrow{d} \\
    0 \arrow[dashed]{r} & \ker \TrFrob{\pi_0(G)} \arrow[dashed]{r} \dar & \arrow{d} \ker \TrFrob{G} \arrow[dashed]{r} & \ker \TrFrob{G^0} =0 \arrow[dotted, out=-10, in=170]{dddll} \dar & \\
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar \dar
    & \arrow{d} G(\Fq)^* \rar & G^0(\Fq)^* \rar \dar & 0\\
    & \arrow{d} \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \arrow{d} \coker \TrFrob{G} \arrow[dashed]{r} &  \arrow{d} \coker \TrFrob{G^0} =0 & \\
      & 0 & 0 & 0 \\
  \end{tikzcd}
  \]
  be the commutative diagram of abelian groups obtained by applying
  Proposition~\ref{prop:pullback} to \eqref{eq:pi0}.
 %
  The sequence of abelian groups
  \begin{equation}\label{eq:pi0k}
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0,
  \end{tikzcd}
  \end{equation}
  is exact since $\Hh^1(\Fq,G^0) =0$ by Lemma~\ref{lem:G0alg-grp} and Lang's theorem on connected algebraic groups over finite fields \cite{lang:56a}.
  Since $\EEx$ is divisible, $\Hom_\text{Ab}(\ - \ ,\EEx)$ is exact and thus the dual sequence of
  character groups is also exact.
%
  The upper row is exact by Proposition~\ref{prop:middleexact}.
  Now Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}
  imply that $\ker \TrFrob{G^0} =0$ and $\coker \TrFrob{G^0}=0$,
  while Proposition~\ref{prop:etale-iso} gives $\ker \TrFrob{\pi_0(G)} \iso \Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$
  and $\coker \TrFrob{\pi_0(G)}=0$.
%
It follows from the snake lemma that $\coker \TrFrob{G} =0$
and $\ker \TrFrob{\pi_0(G)} \to \ker \TrFrob{G}$ is an isomorphism.
This calculates $\ker  \TrFrob{G}$ and shows that
\[
  \begin{tikzcd}
0 \arrow{r} & \Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx)) \arrow{r} & \QCiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0 
  \end{tikzcd}
\]
is exact.
\end{proof}

\todo{Comment, in the introduction, that $\Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$ is trivial if $\pi_0({\bar G})$ is finite cyclic. Lusztig.}

@@

\begin{proposition}\label{prop:basechange}
With notation above,
the rule $\nu_{k'/k}: (\gqcs{L}, \mu, \phi) \mapsto (\gqcs{L}, \mu, \phi_{k'})$
 defines a monoidal functor $\QC(G) \to \QC(G_{k'})$
 such that the following diagram commutes:
\[
\begin{tikzcd}[column sep=60]
\QCiso{G} \rar{\nu_{k'/k}} \dar{\TrFrob{G}} & \QCiso{G_{k'}} \dar{\TrFrob{G_{k'}}} \\
G(k)^*  \rar{\Nm^*} & G(k')^*.
\end{tikzcd}
\]
\end{proposition}

@@

Finally, let $G'$ be a smooth commutative group scheme over $k'$.
We explain how to geometrize the canonical isomorphism between characters of $G'(k')$ and of $(\Res_{k'/k}G')(k)$.
We may decompose the base change $(\Res_{k'/k}G')_{k'}$ of $\Res_{k'/k}G'$ to $k'$
into a product of copies of $G'$, indexed by elements of $\Gal(k'/k)$:
\[
(\Res_{k'/k}G')_{k'} \cong \prod_{\Gal(k'/k)} G'.
\]
Since products and coproducts agree for group schemes we have a natural inclusion of $k'$-schemes
\[
G' \hookrightarrow (\Res_{k'/k}G')_{k'}
\]
mapping $G'$ into the summand corresponding to $1 \in \Gal(k'/k)$.  Composing $\nu_{k'/k}$
from Proposition~\ref{prop:basechange} with pullback along this map yields a functor
\[
\rho : \QC(\Res_{k'/k}G') \to \QC(G').
\]

\begin{proposition}
Let $k'/k$ be a finite extension and let $G'$ be a smooth commutative group scheme over $k'$.
Then the functor 
\[
\rho : \QC(\Res_{k'/k}G') \to \QC(G'),
\]
defined above, induces
\[
\begin{tikzcd}
\QCiso{\Res_{k'/k} G'} \dar{\TrFrob{\Res_{k'/k} G'}} \rar{\rho} & \QCiso{G'} \dar{\TrFrob{G'}}\\
G'(k')^* \rar & G'(k')^*,
\end{tikzcd}
\]
where the bottom map is the identity.
\end{proposition}
\begin{proof}
By Proposition~\ref{prop:pullback} the pullback part of the definition of $\rho$ corresponds to the map
\[
(\Res_{k'/k}G')(k') \to G'(k')^*
\]
induced by $g \mapsto (g, 1, \ldots, 1)$.  Since the action of $\Gal(k'/k)$ on
\[
(\Res_{k'/k}G')_{k'} \cong \prod_{\Gal(k'/k)} G'
\]
is given by permuting coordinates, composition with the norm map yields the identity on $G'(k')$.
\end{proof}

%\todo{Can we improve on this result by proving that $\rho$ is an equivalence of categories?}

@@

Let $K$ be a local field (in the sense of Serre) with residue field $\Fq$. 
In this section we consider connected commutative algebraic groups over $K$ that admit a Néron model, by which we means a lft Néron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected commutative algebraic groups over $K$ that contain no subgroup of type $\mathbb{G}_\text{a}$.
These algebraic groups over $K$ form an additive category, denoted by $\mathcal{N}_K$ below, in a natural way, as a full subcategory of connected commutative algebraic groups over $K$. This category includes algebraic tori over $K$ and Abelian varieties over $K$, and sometimes unipotent K-wound groups, depending on $K$.

Let $R$ be the ring of integers of $K$.
Write $\m$ for the maximal ideal of $R$ and set $R_n = R/\m^{n+1}$, for every non-negative integer $n$.
Suppose $X_K \in \mathcal{N}_K$ with Néron model $X$ over $R$.
A {\it quasicharacter of $X_K(K)$} is a group homomorphism $X_K(K) \to \EEx$ that factors through $X_K(X) \to X(R_n)$, for some non-negative integer $n$;
the group of quasicharacters of $X_K(K)$ will be denoted by $\Hom_\text{qc}(X_K(K),\EEx)$ and the subgroup of those that factor through $X(R_n)$ will be denoted by $\Hom_n(X_K(K),\EEx)$.
%
In this section we explain how to geometrize and categorify quasicharacters of $X_K(K)$ using quasicharacter sheaves.

@@

Let $K$, $R$ and $R_n$ be as above.
Set $S = \Spec{R}$ and $S_n = \Spec{R_n}$, for every non-negative integer $n$;
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.

We will need the Greenberg functor, which provides
a functor from the category of schemes 
over $R_n$, for each $n\in \NN$, to the category of schemes over $\Fq$;
See \cite{bertrapelle-gonzales:Greenberg} for the definition and fundamental properties of the Greenberg functor as we use it; other important references include
\citelist{
\cite{greenberg:61}, \cite{greenberg:63a},
\cite{demazure-gabriel:GroupesAlgebriques}*{V, \S 4, no. 1},
\cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6},
\cite{nicaise-sebag:motivicSerre}*{\S 2.2}
}.
This functor respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components.
%
For any non-negative integer $n$ we will write
\[
\Gr^R_n : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}
\]
for the functor produced by composing pullback along $S_n \to S$ with the Greenberg functor. 
This functor also respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components, and
\[
\Gr^R_n(X)(\Fq) \iso X(R_n),
\]
canonically, for any $X$ lft over $S$.
